Typed Lua: An Optional Type System for Lua

### 2015-05-22 doc/paper/typedlua.tex
Final version published in DYLA

\documentclass[preprint]{sig-alternate}

@@

\newcommand{\Any}{\mathbf{any}}
\newcommand{\Top}{\mathbf{value}}
\newcommand{\Nil}{\mathbf{nil}}
\newcommand{\False}{\mathbf{false}}
\newcommand{\True}{\mathbf{true}}
\newcommand{\Boolean}{\mathbf{boolean}}
\newcommand{\Number}{\mathbf{number}}
\newcommand{\String}{\mathbf{string}}
\newcommand{\Void}{\mathbf{void}}
\newcommand{\Const}{\mathbf{const}}

@@

\conferenceinfo{Dyla}{'14, Edinburgh, UK}

@@

This paper presents the initial design of Typed Lua, an
optionally-typed extension of the Lua scripting language.
The design of Typed Lua's type system has a novel combination
of features that preserves some of the idioms that Lua
programmers are used to, while bringing static type
safety to them.
\end{abstract}

@@

Figure~\ref{fig:typelang} gives the abstract syntax of Typed Lua
types. Only {\em first-class types} correspond to actual Lua
values; {\em second-class types} correspond to expression lists,
and Typed Lua uses them to type multiple assignment and function
application. Types are ordered by a subtype relationship, where
any first-class type is a subtype of $\Top$.

@@

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
\multicolumn{3}{c}{\textbf{First-class types}} \\
T ::= & \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Top & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; T \cup T & \textit{disjoint union types}\\
& | \; S \rightarrow S & \textit{function types}\\
& | \; \{F, ..., F\} & \textit{table types}\\
& | \; X & \textit{type variables}\\
& | \; \mu X.T & \textit{recursive types}\\
& | \; X_i & \textit{projection types}\\
L ::= & \multicolumn{2}{l}{\False \; | \; \True \; | \; {<}{\it number}{>} \; | \; {<}{\it string}{>} } \\
B ::= & \multicolumn{2}{l}{\Nil \; | \; \Boolean \; | \; \Number \; | \; \String} \\
F ::= & T:T \; | \; \Const \; T:T & \textit{field types} \\
\multicolumn{3}{c}{} \\
\multicolumn{3}{c}{\textbf{Second-class types}} \\
S ::= &  \multicolumn{2}{l}{\; V \; | \; V \cup S} \\
V ::= & \;\; T & \\
& | \; T* & \textit{vararg types}\\
& | \; T \times V & \textit{tuple types}
\end{array}
$$
\dend
\caption{Abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

@@

Typed Lua always adds a
variadic tail to the parts of a function type if none is specified, to match
the semantics of Lua function calls. In the examples above, the types
of {\tt factorial} and {\tt abs} are actually $\Number \times \Top *
 \rightarrow \Number \times \Nil *$, and the type of {\tt distance} is
$\Any \times \Any \times \Top * \rightarrow \Number \times \Nil *$.

@@

@@

Typed Lua lets assignment change the type of a local variable in
cases that respect a simple heuristic: the new type must be a
subtype of the previous one, the variable must be local
to the current function, and the variable must not have been
assigned to in another function. The new type only applies for
the remainder of the current scope. In the case of {\tt
greeting}, the assignment on line three changes its type to
$\String$.

@@

Typed Lua has a small set of type predicates that, when used
over a local variable in a condition, constrain the type of that
variable. The function above uses the {\tt type(X) == "string"}
predicate which constrains the type of $X$ from $T \cup \String$ to
$\String$ when the predicate is true and $T$ otherwise. This is
a simplified form of {\em flow typing}~\cite{guha:flow}.

@@

Currently, a limitation of our overloading mechanisms is that
the return type cannot depend on the input types; we cannot
write a function that is guaranteed to return a number if
passed a number and guaranteed to return a string if passed
a string, for example.

@@

The concrete syntax of Typed Lua has syntax for
common table types. One syntax defines table types for
maps: it is written \texttt{\{ t: u \}},
and maps to the table type $\{t:u\}$.
This table type represents a map that maps values of type
$t$ to values of type $u$.
Another syntax defines table types for arrays:
it is written \texttt{\{ t \}}, and maps to the table type
$\{\Number:t\}$. A third syntax sugar defines
table types for records:
it is written \texttt{\{ s1: t1, ..., sn: tn \}}, where
each $s_i$ is a literal number, string, or boolean, 
and maps to the table type $\{s_{1}:t_{1}, ..., s_{n}:t_{n}\}$,
where each $s_i$ is the corresponding literal type.

@@

@@

Typed Lua tags a variable with a table type as either
{\em open} or {\em closed}. If a variable gets its type
from a table constructor then it is open, otherwise it is
always closed. The type of an open variable may change
by field assignment, subject to two restrictions: the
variable must be local to the current block, the new type
must be a subtype of the old type, and the variable
cannot have been assigned to in another function.

@@

We can use recursive table types to represent objects,
and Typed Lua has syntactic sugar to make defining these
types easier:

@@

The double arrow in the type of the two methods is
syntactic sugar for having a first parameter named
{\tt self} with the same type as the interface. The
{\tt const} qualifier is necessary for covariance
in the types of the methods, and to make subtyping
among object types work.

Type-checking a method call is straightforward, but
there is still the matter of how to construct a value with
the object type above. The following example shows one
way:

@@

The {\tt :} syntactic sugar that the example uses also comes
from Lua, and assigns a function to the field with a first
parameter named {\tt self}, plus any other parameters.

Lua has a mechanism for Self-like delegation of missing
fields in a table. After we do {\tt setmetatable(t1,
{ \_\_index = t2 })}, Lua looks up in {\tt t2} any missing
fields of {\tt t1}. Lua programmers often use this mechanism
to simulate classes, as in the following example:

@@

In the last line of the example, notice how we can refer
to {\tt Shape} in the type annotation, as a shortcut
to the recursive table type that Typed Lua has assigned
to this variable.

Typed Lua assigns the type
$\mathbf{Shape} \times \Number \times \Number \times
\Top* \rightarrow \mathbf{Shape} \times \Nil*$ to
{\tt new}. In a {\tt setmetatable} expression,
if the type $t_1$ of the first operand is open and a supertype
of the type $t_2$ of the second operand's {\_\_index} field,
it changes the type of the first operand to $t_1$, and
it remains open.

As the result of {\tt setmetatable} can be an open
table type, we can simulate single inheritance, and
override methods:

@@

If we erase all type and {\tt const} annotations, the
two examples above are valid Lua code, with the same
semantics as the Typed Lua Code.

@@

\citep{tobin-hochstadt2006ims} shows how programs
in the untyped lambda calculus can be incrementally
translated to the simply typed lambda calculus, using
contracts to guarantee that the untyped part cannot
cause errors in the typed part, as a model on how
scripts in a dynamically typed language can be
incrementally translated to a statically typed
language. This approach has been realized in the
Typed Scheme (later Typed Racket) language~\citet{tobin-hochstadt2008ts}.

@@

The main feature of Typed Racket's type system is
{\em occurrence typing}~\citep{tobin:occur}, where
the type of a variable can be refined through tests
using predicates that check the runtime type of
the variable. As these types of checks are common in
other languages, related systems have
appeared~\citep{guha:flow,winther:flow,pearce:whiley}.

>>>>

\documentclass{sig-alternate}

@@

\newcommand{\Any}{\mathbf{any}}
\newcommand{\Top}{\mathbf{value}}
\newcommand{\Nil}{\mathbf{nil}}
\newcommand{\False}{\mathbf{false}}
\newcommand{\True}{\mathbf{true}}
\newcommand{\Boolean}{\mathbf{boolean}}
\newcommand{\Number}{\mathbf{number}}
\newcommand{\String}{\mathbf{string}}
\newcommand{\Void}{\mathbf{void}}
\newcommand{\Const}{\mathbf{const}}
\newcommand{\Self}{\mathbf{self}}

@@

\permission{Permission to make digital or hard copies of all or part of this work for
   personal or classroom use is granted without fee provided that copies are
   not made or distributed for profit or commercial advantage and that copies
   bear this notice and the full citation on the first page. Copyrights for
   components of this work owned by others than the author(s) must be
   honored. Abstracting with credit is permitted. To copy otherwise, or
   republish, to post on servers or to redistribute to lists, requires prior
   specific permission and/or a fee. Request permissions from
   Permissions@acm.org.}
\conferenceinfo{Dyla}{'14, June 09-11 2014, Edinburgh, United Kingdom} 
\CopyrightYear{is held by the owner/author(s). Publication rights licensed to
ACM.}
\crdata{978-1-4503-2916-3/14/06}

@@

This paper presents the initial design of Typed Lua, an
optionally-typed extension of the Lua scripting language.
Lua is an imperative scripting language with first
class functions and lightweight metaprogramming mechanisms.
The design of Typed Lua's type system has a novel combination
of features that preserves some of the idioms that Lua
programmers are used to, while bringing static type
safety to them. We show how the major features of the
type system type these idioms with some
examples, and discuss some of the design issues we faced.
\end{abstract}

\category{D.3.1}{Programming Languages}{Formal Definitions and Theory}[Semantics]
\category{F.3.3}{Logics and Meanings of Programs}{Studies of Program Constructs}[Type structure]

@@

Figure~\ref{fig:typelang} gives the abstract syntax of Typed Lua
types. Only {\em first-class types} correspond to actual Lua
values; {\em second-class types} correspond to expression lists,
and Typed Lua uses them to type multiple assignment and function
application.

Types are ordered by a subtype relationship, where
any first-class type is a subtype of $\Top$. The rest of the 
subtype relationship is standard: union types are supertypes
of their parts, $\Number$, $\Boolean$, and $\String$ are
supertypes of their respective literal types, function
types are related by contravariance on the input part and
covariance in the output part, table types have width
subtyping, with depth subtyping on $\Const$ fields,
tuple and vararg types are covariant.

@@

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
\multicolumn{3}{c}{\textbf{First-class types}} \\
T ::= & \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Top & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; T \cup T & \textit{disjoint union types}\\
& | \; S \rightarrow S & \textit{function types}\\
& | \; \{F, ..., F\} & \textit{table types}\\
& | \; X & \textit{type variables}\\
& | \; \mu X.T & \textit{recursive types}\\
& | \; X_i & \textit{projection types}\\
L ::= & \multicolumn{2}{l}{\False \; | \; \True \; | \; {<}{\it number}{>} \; | \; {<}{\it string}{>} } \\
B ::= & \multicolumn{2}{l}{\Nil \; | \; \Boolean \; | \; \Number \; | \; \String} \\
F ::= & T:T \; | \; \Const \; T:T & \textit{field types} \\
\multicolumn{3}{c}{} \\
\multicolumn{3}{c}{\textbf{Second-class types}} \\
S ::= &  \multicolumn{2}{l}{\; \Void \; | \; V \; | \; V \cup S} \\
V ::= & \;\; T & \\
& | \; T* & \textit{vararg types}\\
& | \; T \times V & \textit{tuple types}
\end{array}
$$
\dend
\caption{Abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

@@

In its default mode of operation, Typed Lua always adds a
variadic tail to the parts of a function type if none is 
specified, to match the semantics of Lua function calls. In the 
examples above, the types
of {\tt factorial} and {\tt abs} are actually
$\Number \times \Top * \rightarrow \Number \times \Nil *$,
and the type of {\tt distance} is $\Any \times \Any \times \Top * 
\rightarrow \Number \times \Nil *$.

@@

There is an optional stricter mode of operation where
Typed Lua does not give variadic tails to the parts
of a function type unless the programmer explictly
declares it, and so will also check all function calls
for arity mismatch.

@@

Typed Lua lets the assignment $x = x \ \mathbf{or} \ e$ change
the type of $x$ from $t \cup \Nil$ to $t$,
as long as the type of $e$ is a subtype of $t$, $x$ is local
to the current function, and it is not assigned in
another function. The change only affects the type of $x$
in the remainder of the current scope. In the case of {\tt
greeting}, the assignment on line three changes its type to
$\String$.

@@

Typed Lua has a small set of type predicates that, when used
over a local variable in a condition, constrain the type of that
variable. The function above uses the {\tt type(x) == "string"}
predicate which constrains the type of $x$ from $T \cup \String$ to
$\String$ when the predicate is true and $T$ otherwise.
This is a simplified form of {\em flow 
typing}~\cite{guha:flow,tobin:occur}. As with {\tt or}, the
variable must be local to the function, and cannot be assigned
to in another function.

@@

Currently, a limitation of our overloading mechanisms is that
the return type cannot depend on the input types; we cannot
write a function that is guaranteed to return a number if
passed a number and guaranteed to return a string if passed
a string, for example. While intersection types provide
a way to express the type of such a function as $\Number 
\rightarrow \Number \cap \String \rightarrow \String$,
more sophisticated flow typing is needed to actually check
that a function has this type, and we are still working on this
problem.

@@

The concrete syntax of Typed Lua has syntax for
common table types. One syntax defines table types for
maps: it is written \texttt{\{ t: u \}},
and maps to the table type $\{t:u\}$.
This table type represents a map that maps values of type
$t$ to values of type $u$.
Another syntax defines table types for arrays:
it is written \texttt{\{ t \}}, and is equivalent to
the table type
$\{\Number:t\}$. A third syntax defines
table types for records:
it is written \texttt{\{ s1: t1, ..., sn: tn \}}, where
each $s_i$ is a literal number, string, or boolean, 
and maps to the table type $\{s_{1}:t_{1}, ..., s_{n}:t_{n}\}$,
where each $s_i$ is the corresponding literal type.

@@

While this runtime check is an instance where the semantics
of a Typed Lua program deviates from the semantics of
plain Lua, the alternatives would be to make all
arrays and maps have an element type that includes {\tt nil},
and either make the programmer narrow the type with {\tt or}
or an {\tt if} statement, making using the elements more
incovenient, or make {\tt nil} a subtype of every type,
reducing the amount of type safety in the system. Notice 
that this does not change the semantics of dynamically
typed programs, as the runtime checks are only added when
the table has a strict static type.

@@

Typed Lua tags a variable with a table type as either
{\em open} or {\em closed}. If a variable gets its type
from a table constructor then it is open, otherwise it is
always closed. The type of an open variable may change
by field assignment, subject to three restrictions: the
variable must be local to the current block, the new type
must be a subtype of the old type, and the variable
cannot have been assigned to in another function.

An variable with an open table type may be aliased, but
these aliases are not open. Any mutation on these aliases
is not a problem, as the type of the original reference
can only change towards a subtype. For mutable fields
this means that the type of the field cannot change
once it is added to the type of the table.

@@

We can use table types and the {\em self type}
to represent objects,
and Typed Lua has syntactic sugar to make defining these
types easier:

@@

The double arrow in the type of the two methods is
syntactic sugar for having a first parameter named
{\tt self} with type $\Self$. The
{\tt const} qualifier is necessary for covariance
in the types of the methods, and to make subtyping
among object types work.

While {\tt :} is syntactic sugar in plain Lua, Typed Lua
uses it to type-check method calls, to bind any ocurrence
of the $\Self$ type in the type of the method to the
receiver. Indexing a method but not immediately calling
it with the correct receiver is a compile-time error.

There is still the matter of how to construct a value with
the object type above. The following example shows one
way:

@@

The {\tt :} syntactic sugar that the example uses also comes
from Lua, and assigns a function to the field with a first
parameter named {\tt self}, plus any other parameters. Typed
Lua adds the {\tt const} annotation, and gives type $\Self$
to the implicit parameter {\tt self}.

Lua has a mechanism for Self-like (or JavaScript-like)
 delegation of missing
fields in a table. After {\tt setmetatable(t1,
\{ \_\_index = t2 \})}, Lua looks up in {\tt t2} any missing
fields of {\tt t1}. Lua programmers often use this mechanism
to simulate classes, as in the following example:

@@

In the last line of the example, notice how we can refer
to {\tt Shape} in the type annotation, as a shortcut
to the table type that Typed Lua has assigned
to this variable.

Typed Lua assigns the type
$\Self \times \Number \times \Number \times
\Top* \rightarrow \Self \times \Nil*$ to
{\tt new}. In a {\tt setmetatable} expression,
if the type $t_1$ of the first operand is a supertype
of the type $t_2$ of the second operand's {\_\_index} field,
it changes the type of the first operand to $t_2$, 
so the local variable {\tt s} 
has the same type as {\tt self}.

We can also use {\tt setmetatable} to simulate single
inheritance, as {\tt setmetatable} on a table constructor
assigned to a fresh local variable gives this variable
an open type, which lets us add new methods and override
existing ones:

@@

If we erase all type and {\tt const} annotations, the
two examples above are valid Lua code, with the same
semantics as the Typed Lua code.

@@

\citet{tobin-hochstadt2006ims} shows how programs
in the untyped lambda calculus can be incrementally
translated to the simply typed lambda calculus, using
contracts to guarantee that the untyped part cannot
cause errors in the typed part, as a model on how
scripts in a dynamically typed language can be
incrementally translated to a statically typed
language. This approach has been realized in the
Typed Scheme (later Typed Racket) language~\citep{tobin-hochstadt2008ts,tobin:occur}.

@@

The main feature of Typed Racket's type system is
{\em occurrence typing}~\citep{tobin:occur}, where
the type of a variable can be refined through tests
using predicates that check the runtime type of
the variable. As these types of checks are common in
other languages, but ocurrence typing is not sound
in the presence of mutation, related systems have
appeared~\citep{guha:flow,winther:flow,pearce:whiley}.
